\vspace{3cm} \Huge{Optimizing SasView models} \\ \Large{Enhancing performance by rewriting kernels from OpenCL/OpenMP to Futhark}

### 2017-11-06 report/draft.tex
b

\sasmodels has built-in default definitions for \texttt{I(q_x,q_y)} if working
with two-dimensional data, and other functions such as \texttt{form_volume(q)}
for working with voluminous scatterings.

Therefore, it is optional for the developer to define a custom definition
of \texttt{I(q_x, q_y)} or other default functions.
An example of this can be seen in appendix dab or line.

@@

\begin{figure}
  % fuld beskrivelse af ax+b-figur, med parametre a, b, og en Iqxy
  \caption{model for modelling a line}\label{fig:linemodel}
\end{figure}

\subsection{Defining \texttt{I(q)} in Python and OpenCL}
A \sasmodels \iq function can be written in either Python (which will be
executed in a Numpy context at runtime), or it can be written in C, which will
will be inserted into an OpenCL template, so that it can be compiled and 
executed as a kernel in an OpenCL context, preferably on a GPU.

A Python \iq can be seen in figure \ref{fig:linemodel}, whereas both DAB (figure
\ref{fig:dab} and core_shell(figure \ref{fig:core_shell}) is written in C.

But with the arrival of Futhark models in \sasmodels, there is now a third
additional way of writing \sasmodels models.

\subsection{Using }

@@

@@

\section{Implementation}
% current architecture of Sasmodels engine
% description of kernelcl kernels
% (smaller) description of kernelcl kernels
% description of current kernelfut implementation
% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
% under runtime, og prægenererede moduler kan deles med andre, uden disse
% ``andre''' behøver have futhark-oversætteren selv.

@@

% discussion of futhark vs. opencl performance
%% where does the performance gains come from?
%%% MASSER AF REFERENCER TIL TROELS' PAPERS

>>>>

\sasmodels has built-in default definitions for $I(q_x,q_y)$ if working
with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
for working with voluminous scatterings.

Therefore, it is optional for the developer to define a custom definition
of $I(q_x, q_y)$ or other default functions.
An example of this can be seen in appendix dab or line.

@@

\begin{figure}
  % fuld beskrivelse af ax+b-figur, med parametre a, b, og en Iqxy
  \caption{model info file for modelling a line}\label{fig:linemodel}
\end{figure}

\subsection{Defining \iq in Python and OpenCL}
A \sasmodels \iq function can be written in either Python (which will be
executed in a Numpy context at runtime), or it can be written in C, which will
will be inserted into an OpenCL template, so that it can be compiled and 
executed as a kernel in an OpenCL context, preferably on a GPU.

A Python \iq can be seen in figure \ref{fig:linemodel}, whereas both DAB (figure
\ref{fig:dab} and core\_shell(figure \ref{fig:core-shell}) is written in C.
Notice that the main difference between the models lies in the Iq property.
For Python models, \iq is a callable Python function, whilst the C \iq is a
 string which is copy-pasted wholesale into aforementioned OpenCL template.

But with the addition of Futhark models in \sasmodels adds a third way of
using and defining a \sasmodels model.

\subsection{Using a Futhark model}
\label{sec:using-futhark}
There is a difference between defining and using a Futhark model.
The \sasview user who wishes to use a Futhark model in his or her work does 
not need more than the futhark model info file, and a copy of the compiled
Futhark kernel. In fact, the futhark model info file is identical to OpenCL- 
and Python model info files, except for in two properties.
As show in figure \ref{fig:futhark_dab}, there has been added a flag 
%CHECK THISSS!!!
\texttt{model\_is\_futhark: true}, and \iq is now a path to a precompiled 
Futhark kernel.

The user does not need to have any Futhark compiler installed if he merely wants
to use the precompiled Futhark module: Futhark models for \sasmodels compiles to
portable Python libraries which has numpy and pyopencl as their only 
dependencies - and as \sasmodels has these dependencies as well, the user will
already have these dependencies sorted out, if he has a working installation of 
\sasview or \sasmodels.

\subsection{Defining a Futhark model}
To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
A model \textit{m} itself is defined as a parametric Futhark module\footnote{
  A more in-depth explanation of the complete definition of a Futhark module is 
  in \ref{sec:generating-futhark}},
which contains an Iq-function, an Iqxy- and a form\_volume function.
It is the model programmers responsibility to fill in all three functions.

%% figure

A simple example of a Futhark model is the line module, 
as depicted in figure \ref{fig:linemodel-futhark}.
Note that Iqxy is non-default, whilst form\_volume is the default 
implementation.

\subsubsection{On the user friendliness of Futhark models}
As shown in figure \ref{fig:linemodel-futhark}, there are currently two small 
nuisances in Futhark models, compared to equivalent models written in
Python/OpenCL:

With both Python and OpenCL models, the programmer can leave out Iqxy and 
form\_volume so that \sasmodels automatically fills in the default definition
for these two functions instead.

The Futhark programmer does not have this leisure yet, and must define the two
functions no matter what. This is admittedly a piece of annoying boiler plate,
but it is also a thing that can be solved by compartmentalising the offending
functions away in a parametric library module.

The less removable nuisance is that the programmer must explicitly extract the
Iq function parameters from the list \texttt{local\_values}. This must be done
in the order they appear in the model info parameter list.

Iq functions in Python and OpenCL can be defined with any number of function 
parameters, but in Futhark, we need to adhere to the type declaration of Iq 
(see \ref{has-iq}), which means that Iq can take only two arguments; the 
scattering vector \texttt{q}, and the list \texttt{local\_values} of parameters
needed for the scattering.

This makes for an ugly prelude to the actual body of the Iq function, as shown
in fig\ref{fig:linemodel-futhark}, or more critically in models like 
core-shell \ref{fig:core-shell}.

\subsection{Generating Futhark models using parametric modules and a header}
\label{sec:generating-futhark}
\begin{lstlisting}
module type has_iq = {
Iq: real -> real[] -> real // the model's scattering function.
Iqxy: real -> real -> real[] -> real // the model's 2D scattering function
form_volume: real[] -> real // the model's form volume function.
}  
\end{lstlisting}

@@

\subsubsection{Why not Futhark models as model info strings instead standalone modules?}
% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
% under runtime, og prægenererede moduler kan deles med andre, uden disse
% ``andre''' behøver have futhark-oversætteren selv.

@@

\section{Implementation}
% current architecture of Sasmodels engine
% description of kernelcl kernels
% (smaller) description of kernelcl kernels
% description of current kernelfut implementation

@@

% discussion of futhark vs. opencl performance
%% where does the performance gains come from?
% show core_shell-whatever; and elaborate on loop flattening into maps
%%% MASSER AF REFERENCER TIL TROELS' PAPERS
